---
title: "R Basics Practice"
output: html_document
---

## Step 1: Read in the data using `read_csv()`

Save it as a new data frame called `yaounde_covid_seroprev`.

The dataset is saved online, not locally in this folder. Its location is "https://raw.githubusercontent.com/kendavidn/yaounde_serocovpop_shared/v1.0.0/data/yaounde_covid_seroprev_dataset.csv"

HINT: Remember to load the `tidyverse` package before trying to use the `read_csv()` command.

```{r}
library(tidyverse)

yaounde_covid_seroprev <- read_csv("https://raw.githubusercontent.com/kendavidn/yaounde_serocovpop_shared/v1.0.0/data/yaounde_covid_seroprev_dataset.csv")
```

## Step 2: Inspect the data using the Data Viewer.

Open the `yaounde_covid_seroprev` data in the Data Viewer.
Answer the following questions:

- How many observations (rows) are in the data? How many columns?

There are 1011 rows and 303 columns.

- Sort the data by `val_age`. How old is the youngest person in the data? How old is the oldest?

The youngest person is 5, the oldest is 79.

HINT: Can't find the `val_age` column? Note that RStudio only shows 50 columns at a time, so you may need to click over to the next 50 columns (columns 51-100) to find what you're looking for. Look for where it says "Cols: 1-50" at the top.

## Step 3: Use dplyr to save a version of the data that will be easier to work with

This dataset includes many columns that we don't actually need for our current analysis. 
Use dplyr to save a new version of the data called `data_cleaned` that selects only the following columns: `val_age`, `had_symp_none`, `cat_igm_result`, and `cat_igg_result`. 

```{r}
data_cleaned <- select(yaounde_covid_seroprev, val_age, had_symp_none, cat_igm_result, cat_igg_result)

# or, with a pipe (this code will do exactly the same thing):
data_cleaned <- yaounde_covid_seroprev %>% 
  select(val_age, had_symp_none, cat_igm_result, cat_igg_result)
```


## Step 4: Inspect the new data in the Data Viewer

Open `data_cleaned` in the Data Viewer, and answer the following questions:

- Filter the data by `had_symp_none` = 0. How many observations reported any symptoms?

There are 303 observations after filtering for `had_symp_none` = 0.

- Remove the filter on `had_symp_none`. Filter the data by `cat_igg_result` = "positif". How many total observations had a positive igg test?

There are 304 observations with a positive igg test (`cat_igg_result` = "positif").

- Of the observations with any symptoms (`had_symp_none` = 0), how many had a positive igg result? How many had a positive igm result?

Filtering for both `had_symp_none` = 0 and `cat_igg_result` = "positif" gives 108 observations.
Filtering for both `had_symp_none` = 0 and `cat_igm_result` = "positif" gives 12 observations.

## Step 3: Create a histogram

Use ggplot and the `data_cleaned` data frame to create a histogram of `val_age`. Adjust the binwidth until you get something you like.

```{r}
ggplot(data = data_cleaned) +
  geom_histogram(mapping = aes(x = val_age))
```

```{r}
# try a binwidth of 1 year
ggplot(data = data_cleaned) +
  geom_histogram(mapping = aes(x = val_age), binwidth = 1)

# try a binwidth of 10 years
ggplot(data = data_cleaned) +
  geom_histogram(mapping = aes(x = val_age), binwidth = 10)
```


Next, map a fill aesthetic to the histogram. Color the bars by `cat_igg_result` by using the `fill` aesthetic.

```{r}
ggplot(data = data_cleaned) +
  geom_histogram(mapping = aes(x = val_age, fill = cat_igg_result), binwidth = 5)
```

## Step 4: Create a boxplot

Try a different geom. Make a boxplot of the same data, `val_age`.

```{r}
ggplot(data = data_cleaned) +
  geom_boxplot(mapping = aes(x = val_age))
```

```{r}
# here it is with the fill aesthetic added as well
ggplot(data = data_cleaned) +
  geom_boxplot(mapping = aes(x = val_age, fill = cat_igg_result))
```

```{r}
# note that if you want to, you can specify y = val_age instead of x = val_age
ggplot(data = data_cleaned) +
  geom_boxplot(mapping = aes(y = val_age, fill = cat_igg_result))
```

## Step 5: Create a dplyr pipeline to save just the asymptomatic data

Let's say we want to look specifically at asymptomatic cases.

Create a dplyr pipeline that:

1. Starts with the *yaounde_covid_seroprev* data frame, then

2. Filters to observations that had no symptoms (Hint: `had_symp_none` equal to 1), then

3. Uses `mutate` to create a new column called `log_igg_result` (we're using "log" as short for "logical" here) that is the result of a logical test of whether `cat_igg_result` equals "positif". Do the same for igm results (create a new column called `log_igm_result` that is the result of a logical test of whether `cat_igg_result` equals "positif"). Create a third new column called `log_result` that shows whether the observation had either a positive igg or igm test (HINT: you'll need a logical statement that tests for "`log_igg_result` or `log_igm_result`"). Then,

4. Selects only the columns you want to work with: `val_age`, `cat_igg_result`, `cat_igm_result`, and `log_result`.

5. Saves that as a new dataframe called `asymp_data`.

HINT: Stuck on #3 above? Remember that mutate commands in a pipeline (where the data is being piped in) take the form `mutate(new_col = definition)` where `new_col` is the name of the column you want to create and `definition` is how that column should be specified. In this case, the definition will be a logical statement (remember that logical tests of equality in R need `==`!). A single vertical line `|` is the logical symbol for "or". 

```{r}
asymp_data <- yaounde_covid_seroprev |> 
  filter(had_symp_none == 1) |> 
  mutate(log_igg_result = cat_igg_result == "positif",
         log_igm_result = cat_igm_result == "positif",
         log_result = log_igg_result | log_igm_result) |> 
  select(val_age, cat_igg_result, cat_igm_result, log_result)
```


## Step 6 Create a histogram using the asymptomatic data

1. Create a histogram for `val_age` using the `asymp_data`. 

2. Add a `ggtitle()` layer to make the title "COVID-19 IGG/IGM Testing" with the subtitle "Asymptomatic Cases Only".

3. Add a `xlab()` layer to make the x-axis label "Age (years)"

3. Change the theme using `theme_bw()` (or try out a different theme, if you like!)

4. Add a `facet_wrap()` layer to split it into separate histograms based on the `log_result` column.

```{r}
ggplot(data = asymp_data) +
  geom_histogram(mapping = aes(x = val_age), binwidth = 5) + 
  ggtitle(label = "COVID-19 IGG/IGM Testing", 
          subtitle = "Asymptomatic Cases Only") + 
  xlab("Age (years)") +
  theme_bw() + 
  facet_wrap(~ log_result)
```

Well done! That's the end of the practice problems. Return to the module to read the last sections. 

