---
title: "Visualize Data"
output: html_document
---

### Your Turn: The ggplot template and aesthetics

Run the following code chunk to import the COVID19 data.

```{r, warning = FALSE, message = FALSE}
library(tidyverse)

covid_testing <- read_csv("covid_testing.csv")
```



Let's recreate the histogram of *pan_day*.

You will need to:

1. Pick a data set: *covid_testing*
2. Choose a geom function: *geom_histogram()*
3. Map aesthetic properties of the graph to variables of the data frame (in this case, the *pan_day* variable should be on the x axis).

Fill in the blanks and run the code to create the plot.

```{r}
ggplot(data = covid_testing) +
  geom_histogram(mapping = aes(x = pan_day))
```




When you run this code, you will most likely have what looks like an error displayed above the graph:

  `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.

This is actually just a message from R, telling you that it's a bad idea to use a default bin width.

The *pan_day* variable holds the day of the pandemic (an arbitrary day in late February of 2020) on which the COVID test order was placed. What would be an appropriate binwidth for *pan_day* to see *daily* test volumes?

Now fill in the blanks below to recreate the code from the previous chunk, with the *binwidth* argument passed to the *geom_histogram* function.

```{r}
ggplot(data = covid_testing) +
  geom_histogram(mapping = aes(x = pan_day), binwidth = 1)
```




The effect of reducing the bin width (and increasing the number of bins) is that the data is now presented in much greater detail.

Let's next add some color!  This time we'll use "fill" to map the fill color.  What you're trying to accomplish is change the interior (or "fill") color to correspond to a category (`result`).  You're going to have *two* arguments inside `aes` now.

How to do that? You can put several `aes` arguments in a row, separated by a comma, like this (obviously a silly example):

`aes(x = something, color = something_else, fill = some_third_thing)`

Copy and paste the previous code chunk and add an aesthetic mapping that maps the *fill* aesthetic to the *result* column. (Hint: all aesthetics must be defined *inside* the `aes()` function!)

```{r}
ggplot(data = covid_testing) +
  geom_histogram(mapping = aes(x = pan_day, fill = result), binwidth = 1)
```







***************************************************************************************
***************************************************************************************
***************************************************************************************
***************************************************************************************
***************************************************************************************
********************************                    ***********************************
********************************     Stop here.     ***********************************
********************************                    ***********************************
***************************************************************************************
***************************************************************************************
***************************************************************************************
***************************************************************************************
***************************************************************************************






### Your Turn: Geom functions

Run the following code chunk:

```{r}
ggplot(data = covid_testing) +
  geom_histogram(mapping = aes(x = pan_day))
```


Now try to figure out how you would modify the code above so it draws a frequency polygon instead of histogram.

```{r}
ggplot(data = covid_testing) +
  geom_freqpoly(mapping = aes(x = pan_day))
```




Modify the previous code chunk so that the frequency polygon's line color is blue. (Hint: color names go in quotes!)

```{r}
ggplot(data = covid_testing) +
  geom_freqpoly(mapping = aes(x = pan_day), color = "blue")
```







What do you think the following code will do? Try to predict what you'll see.

Then run it.

```{r}
ggplot(data = covid_testing) +
  geom_histogram(mapping = aes(x = pan_day)) +
  geom_freqpoly(mapping = aes(x = pan_day))
```

## More Visualization Options

### Position Adjustments

```{r}
# instead of a stacked histogram, you can request side-by-side bars
ggplot(covid_testing) +
  geom_histogram(
    mapping = aes(x = pan_day, fill = result),
    position = position_dodge()
  )
```

### Themes

```{r}
# use different themes which affect how non-data elements such as axes, gridlines, and background appear

ggplot(covid_testing) +
  geom_histogram(
    mapping = aes(x = pan_day, fill = result),
    position = position_dodge()
  ) +
  theme_light()
```

### Scales

```{r}
# customize color scales

library(colorspace)

cols <- c(
  "grey80",
  qualitative_hcl(2, palette = "dark3")
)
names(cols) <- c("invalid", "negative", "positive")

ggplot(covid_testing) +
  geom_histogram(
    mapping = aes(x = pan_day, fill = result),
    position = position_dodge()
  ) +
  theme_light() +
  scale_fill_manual(values = cols)
```

### Facets

```{r}
# break a plot into sub-plots by another variable
ggplot(covid_testing) +
  geom_histogram(
    mapping = aes(x = pan_day, fill = result)
  ) +
  theme_light() +
  scale_fill_manual(values = cols) +
  facet_wrap(~demo_group)
```

### Titles and captions

```{r}
# add titles, subtitles, or annotations, and change the axis labels or the appearance
ggplot(covid_testing) +
    geom_histogram(
        mapping = aes(x = pan_day, fill = result)
    ) +
    theme_light() +
    facet_wrap(~demo_group) +
    ggtitle(label = "COVID19 Test Volume",
         subtitle = "Faceted by Demographic Group") +
    xlab("Day of Pandemic") +
    ylab("Number of Tests")
```
